\textsf{\textbf{Analyzing \textit{Streptococcus mutans} RNA-Seq data using R packages}}

### 2012-01-06 package/smutans/inst/doc/smutans.Rnw
A smutans R package version of 0.1.2 is used. I still need to work on UA159/TW1 comparisons.

We can create \Rclass{CountDataSet} objects using functions that are not
exported. One should create an R script at the directory that contains the
source code of the \Rpackage{smutans}. Put the following scripts in the R script
and run the script at the directory. That should create RData files at
the directory smutans/data: sm835Genes.RData, smutansGenes.RData.
<<samplePreparation,eval=FALSE>>=
#not run
library(DESeq)
source("smutans/R/core.R")
smutans.prepareData835NPP( )
smutans.prepareDataUA159TW1( ) 
q("no")
@

@@

We must have at least one citation \cite{Anders:2010:GB}.

@@

\begin{figure}
\centering
\includegraphics[width=.9\textwidth]{smutans-figClust}
\caption{Heatmap showing the Euclidean distances between the samples
  as calculated from the variance-stabilising transformation of the
  count data.}
\label{figClust}
\end{figure}

<<figClust,fig=TRUE>>=
smutans.de2Clust( allSample )
@

@@

@@

<<ua159onlySamples>>=
npAndP <- newSmutans( sm835Genes, title="835NP vs. 835P" )
npAndP <- smutans.de2( npAndP, type="ua159", 
                       condA="835np", condB="835p" )
@

\begin{figure}
\centering
\includegraphics[width=.6\textwidth]{smutans-figFitNpAndP}
\caption{Empirical (black dots) and fitted (red lines) dispersion values
plotted against mean expression strength. \textbf{figFitNpAndP}}
\label{figFitNpAndP}
\end{figure}
<<figFitNpAndP,fig=TRUE>>=
smutans.plotDispersionEstimates(npAndP)
@

@@

\bibliographystyle{unsrt}
\bibliography{smutans}

>>>>

We can create \Rclass{CountDataSet} objects using functions that are not
exported. One should create an R script at the directory that contains the
source code of the \Rpackage{smutans}. Put the following scripts in the R script
and run the script at the directory. That should create RData files at
the directory smutans/data: sm835Genes.RData, smutansGenes.RData.
<<samplePreparation,eval=FALSE>>=
#not run
library(DESeq)
source("smutans/R/core.R")
smutans.prepareData835NPP( )
smutans.prepareDataUA159TW1( ) 
smutans.prepareGoseq( )
q("no")
@

@@

% We must have at least one citation \cite{Anders:2010:GB}.

@@

%\begin{figure}
%\centering
%\includegraphics[width=.9\textwidth]{smutans-figClust}
%\caption{Heatmap showing the Euclidean distances between the samples
%  as calculated from the variance-stabilising transformation of the
%  count data.}
%\label{figClust}
%\end{figure}

<<figClust,fig=TRUE>>=
# smutans.de2Clust( allSample )
@

We find some functional categories enriched with differentially expressed genes.
<<goseqUA159only>>=
data( "smutans.feature.genes" )
data( "smutans.go.genes" )
data( "smutans.cat.desc" )
smutans.de2Goseq ( ua159only, qval=1e-2, 
                   feature.gene=smutans.feature.genes,
                   go.genes=smutans.go.genes,
                   cat.desc=smutans.cat.desc )
@

@@

A count data set is a table of numbers of short reads that are assigned to
predefined genes. We use one of R packages for testing for differential
expression: \Rpackage{DESeq}.

To find functional categories enriched with differentially expressed genes we
first find a list of differentially expressed genes. A set of genes are assigned
to functional gene categories: \Robject{smutans.go.genes}. Features of the genes in
UA159 include gene lengths: \Robject{smutans.feature.genes}.
Functional categories with numbers of genes assigned are in
\Robject{smutans.cat.desc}.

We use RNA-seq data to define genes whose expression changes between
samples \cite{Oshlack2010}. Short reads are first mapped to a
reference genome using approaches such as BWA \cite{Li2009a}.
Reads fitting into one of the annotated intervals are counted.
These counts are passed to a statistical package which treats differences
in read counts for statistical significance. For this purpose, we made
use of DEseq \cite{Anders2010}. This prodecure returns a list of
differentially expressed genes, which can be passed to downstream
analysis, such as functionaly category enrichment test (e.g., Young et
al., 2010).

The abtract briefly describes main steps of analyzing RNA-seq data for
determining differentially expressed genes \cite{Oshlack2010}. We
used a short read aligner called BWA \cite{Li2009a} to map the
sequenced short reads on a reference genome, \textit{Streptococcus mutans}
UA159 (NCBI accession NC\_004350). Because the aligner allowed a few gaps
for efficient alignment of millions of sequences of size being 100 base
pairs long, short reads that consist of RNAs and sequencing adapter
sequences would be not mapped. We used cutadapt \cite{Martin2011} to remove
parts of adapter sequences from short reads before mapping on the reference
genome. Mapped short read alignments were converted into readable formats
\cite{Li2009b}. We counted short reads aligned to annotated genes in the
reference genome, forming a table of read counts of the genes. Statistical
software R \cite{R2011} with its one of R package
called DEseq \cite{Anders2010} were subsequently employed to infer
significantly differentially expressed in a pair of biological conditions.
We also generated new tracks for our recently released Streptococcus Genome
Browser (http://strep-genome.bscb.cornell.edu) that summarize gene
expression levels based the short read alignments. This allowed us to
visualize parts of the reference genome for comparing different RNA
samples. Further analyses such as functionaly category enrichment test
\cite{Young2010} were pursued.

We could find no functional categories enriched with differentially expressed
genes in the pairwise comparisons of 835NP, 835P, and UA159 no CSP.

Let's make a list of differentially expressed genes. It could be interesting to
see the cluster of the three kinds of samples. Another kind of heatmap plots for
grouping genes could be interesting. But, basically we have a list of genes with
some adjusted p-values for detecting differentially expressed genes.

@@

<<sm835Samples>>=
npAndP <- newSmutans( sm835Genes, title="835NP vs. 835P" )
npAndP <- smutans.de2( npAndP, type="ua159", 
                       condA="835np", condB="835p" )
npAndUA159 <- newSmutans( sm835Genes, title="835NP vs. UA159" )
npAndUA159 <- smutans.de2( npAndUA159, type="ua159", 
                           condA="835np", condB="ua159nocsp" )
pAndUA159 <- newSmutans( sm835Genes, title="835P vs. UA159" )
pAndUA159 <- smutans.de2( pAndUA159, type="ua159", 
                          condA="835p", condB="ua159nocsp" )
@

\begin{figure}
\centering
\includegraphics[width=.6\textwidth]{smutans-figFitNpAndP}
\includegraphics[width=.6\textwidth]{smutans-figFitNpAndUA159}
\includegraphics[width=.6\textwidth]{smutans-figFitPAndUA159}
\caption{Empirical (black dots) and fitted (red lines) dispersion values
plotted against mean expression strength. \textbf{figFitNpAndP}}
\label{figFitNpAndP}
\end{figure}
<<figFitNpAndP,fig=TRUE>>=
smutans.plotDispersionEstimates(npAndP)
@
<<figFitNpAndUA159,fig=TRUE>>=
smutans.plotDispersionEstimates(npAndUA159)
@
<<figFitPAndUA159,fig=TRUE>>=
smutans.plotDispersionEstimates(pAndUA159)
@

<<sm835Samples>>=
smutans.de2List( npAndP, "npAndP.csv" )
smutans.de2List( npAndUA159, "npAndUA159.csv" )
smutans.de2List( pAndUA159, "pAndUA159.csv" )
@

%\begin{figure}
%\centering
%\includegraphics[width=.9\textwidth]{smutans-figClustNpAndP}
%\caption{Empirical (black dots) and fitted (red lines) dispersion values
%plotted against mean expression strength. \textbf{figClustNpAndP}}
%\label{figClustNpAndP}
%\end{figure}
<<figClustNpAndP,fig=TRUE>>=
# smutans.de2Clust( npAndP )
@
%\begin{figure}
%\centering
%\includegraphics[width=.9\textwidth]{smutans-figClustNpAndUA159}
%\caption{Empirical (black dots) and fitted (red lines) dispersion values
%plotted against mean expression strength. \textbf{figClustNpAndUA159}}
%\label{figClustNpAndUA159}
%\end{figure}
<<figClustNpAndUA159,fig=TRUE>>=
# smutans.de2Clust( npAndUA159 )
@
%\begin{figure}
%\centering
%\includegraphics[width=.9\textwidth]{smutans-figClustPAndUA159}
%\caption{Empirical (black dots) and fitted (red lines) dispersion values
%plotted against mean expression strength. \textbf{figClustPAndUA159}}
%\label{figClustPAndUA159}
%\end{figure}
<<figClustPAndUA159,fig=TRUE>>=
# smutans.de2Clust( pAndUA159 )
@

\begin{figure}
\centering
\includegraphics[width=.9\textwidth]{smutans-figHeatmapNpAndP}
\caption{Empirical (black dots) and fitted (red lines) dispersion values
plotted against mean expression strength. \textbf{figHeatmapNpAndP}}
\label{figHeatmapNpAndP}
\end{figure}
<<figHeatmapNpAndP,fig=TRUE>>=
smutans.de2TypeHeatmap( npAndP )
@
\begin{figure}
\centering
\includegraphics[width=.6\textwidth]{smutans-figHeatmapNpAndUA159}
\caption{Empirical (black dots) and fitted (red lines) dispersion values
plotted against mean expression strength. \textbf{figHeatmapNpAndUA159}}
\label{figHeatmapNpAndUA159}
\end{figure}
<<figHeatmapNpAndUA159,fig=TRUE>>=
smutans.de2TypeHeatmap( npAndUA159 )
@
\begin{figure}
\centering
\includegraphics[width=.6\textwidth]{smutans-figHeatmapPAndUA159}
\caption{Empirical (black dots) and fitted (red lines) dispersion values
plotted against mean expression strength. \textbf{figHeatmapPAndUA159}}
\label{figHeatmapPAndUA159}
\end{figure}
<<figHeatmapPAndUA159,fig=TRUE>>=
smutans.de2TypeHeatmap( pAndUA159 )
@

@@

\bibliographystyle{unsrt}
\bibliography{siepel-mutans}
